---
title: "3D Partial Dependence Plots (PDPs) using the plotly and pdp packages"
author: "STATS-ML"
date: "9/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A function to create a contour PDP plot for 2 variables
```{r include = TRUE}
# A function to return a partial dependence contour plot
pdp_contour <- function(var1, var2, optimal_number_of_trees, model_object){
  # var1 (string): name of one of the two predictors used in the plot
  # var2 (string): name of one of the two predictors used in the plot
  # optimal_number_of_trees (integer): optimal number of trees based on cross-validation or a testing sample 
  # model_object (gbm): the gbm model used to construct the PDP
  
  # Generate the partial dependence given the variables
  pd <- model_object %>% partial(pred.var = c(var1, var2), n.trees = optimal_number_of_trees)
  # Interpolate the partial dependence values to generate a surface 
  dens <- akima::interp(x = pd[,var1], y = pd[,var2], z = pd$yhat)
  
  # Flattened contour partial dependence plot for 2 variables
  p <- plot_ly(x = dens$x, y = dens$y, z = dens$z, colors = c("blue", "grey", "red"), type = "contour")
  # Add axis labels for 2D plots
  p <- p %>% layout(xaxis = list(title = var1), yaxis = list(title = var2))
  # Return the plot
  return(p)
}
```

## A funciton to create an interactive 3D Partial Dependency Plot
```{r include = TRUE}
# A function to return a 3D partial dependence plot 
pdp_3D <- function(var1, var2, optimal_number_of_trees, model_object){
  # var1 (string): name of one of the two predictors used in the plot
  # var2 (string): name of one of the two predictors used in the plot
  # optimal_number_of_trees (integer): optimal number of trees based on cross-validation or a testing sample
  # model_object (gbm): the gbm model used to construct the PDP
  
  pd <- model_object %>% partial(pred.var = c(var1, var2), n.trees = optimal_number_of_trees)
  dens <- akima::interp(x = pd[,var1], y = pd[,var2], z = pd$yhat)
  
  # 3D partial dependence plot with a coloring scale
  p <- plot_ly(x = dens$x, y = dens$y, z = dens$z, colors = c("blue", "grey", "red"), type = "surface")
  
  # Add axis labels for 3D plots
  p <- p %>% layout(scene = list(xaxis = list(title = var1),
                                   yaxis = list(title = var2),
                                   zaxis = list(title = "Partial Dependence")))
  # Return the plot
  return(p)
}
```

## Generate Data
```{r include = TRUE}
# Choose a random seed to reproduce the results
set.seed(007)
# Generate 5 variables from a multivariate normal distribution
X <- MASS::mvrnorm(n = 100, mu = rep(0, 5), Sigma = diag(5))
# Generate a binary variable y based on a simple function
y <- ifelse(X[,1] <= 2.1 & X[,2] + X[,3] >= -1,1,0)
# Put the data into a dataframe
data <- data.frame(cbind(y,X))
# Rename the columns "y", "X1", "X2", "X3", "X4", and "X5"
colnames(data) <- c("y", paste("X",1:5,sep=""))
# Show the first 5 rows of the dataframe
head(data)
```

## Construct a GBM model
```{r}
library(gbm)
gbm1 <- gbm(data = data, 
            formula = y ~., 
            distribution = "bernoulli", 
            n.trees = 200, 
            interaction.depth = 3,
            shrinkage = 0.10, 
            bag.fraction = 0.5, 
            cv.folds = 10)

# The optimal number of trees is the number of trees that minimize the cross-validation error
opt_trees <- which.min(gbm1$cv_error)
```

## Interactive 3D Partial Dependence Plot with plotly
#### For binary classification, the veritical axis is the 1/2 log-odds of y = 1 given the predictor (see Friedman 2001)
```{r include = TRUE, message = FALSE, warning = FALSE}
# An interactive 3D PDP for two variables using plotly
# Simply click and drag to rotate the surface (in the IDE not the web browser)
library(plotly)
library(pdp)
pdp_3D(var1 = "X1", var2 = "X2", optimal_number_of_trees = opt_trees, model_object = gbm1)
```

## An interactive flattened PDP (i.e. contour plot) for two variables (easier to understand sometimes)
```{r include = TRUE, message = FALSE, warning = FALSE}
# Hover the mouse over the plot to see the plot values (in the IDE not the web browser)
pdp_contour(var1 = "X1", var2 = "X2", optimal_number_of_trees = opt_trees, model_object = gbm1)
```

